From Questions to Queries – Enhancing the Curt System by a Theory and Implementation of Embedded WH-Questions

### 2008-08-30 thesis/paper.tex
Cleaned up sec 4, added grammar stuff.

@@

\usepackage{fontspec}
\usepackage[xetex,colorlinks]{hyperref}
\begin{document}
\renewcommand{\thepage}{\roman{page}}
\maketitle
\thispagestyle{empty}
\begin{center}
  \vskip 4em
  Thesis submitted for the Degree of Bachelor of Arts\\

@@

\subsection{Embedded Questions}

@@

\begin{definition}
  Resolutive and factive predicates:
  \begin{itemize}
  \item \term{Factive predicates} take both, questions and propositions, and
  demand the extension of (the set of) propositions they embed to evaluate to true (exhaustively).
  Examples:
  \example{know}, \example{forget}, \example{reveal}, \example{discover},
  \item \term{Resolutive predicates} take both, questions and propositions, and
  do not demand the extension of (set of propositions) they embed to have any particular truth value.
  Examples:
  \example{report}, \example{predict}
  \end{itemize}
\end{definition}

@@

TODO Discussion of Lewis' paper.

@@

\subsubsection{Advertent Curt}

@@

\subsubsection{The Syntax of Questions}

@@

However, the syntax of embedded interrogatives is a more delicate matter.

The underlying theory for

\subsubsection{Defining the Grammar Rules}

\curt{} parses sentences of natural language using an enhanced version of \pn{Cooper
storage}\footnote{See \cite{cooper:storage2} for the original implementation.} as
presented in \cite{keller:storage}. \cite{blackburnbos:cl1} call this version
\pn{Keller Storage}. That name will be used from here % FIXME sucks. Please change
to refer to this particular implementation as presented in
\cite{blackburnbos:cl1}.

@@

Enhancing the grammatic rules of the \curt{} system to cope with
proposition-embedding verbs is made trivial by its elaborate grammatical back
end. \ref{ex:kel1} is an example of the logic behind the grammar that is put to
use.

\footnotetext{Note that \ref{ex:kel1} is shown using the same syntactic type system used in
\cite{gs:sawhq}, which extends the system given in \cite{ptq} by $\bar{t}$ to
denote embedded \wh-constituents.}

>>>>

\usepackage{listings}

@@

\usepackage{fontspec}
\usepackage[xetex,colorlinks]{hyperref}
\begin{document}
\lstset{language=Prolog,numbers=left,numberstyle=\tiny,numbersep=5pt,stepnumber=2,basicstyle=\ttfamily,identifierstyle=\textbf,keywordstyle=\underbar}
\renewcommand{\thepage}{\roman{page}}
\maketitle
\thispagestyle{empty}
\begin{center}
  \vskip 4em
  Thesis submitted for the Degree of Bachelor of Arts\\

@@

\subsection{Embedded Questions}\label{sec:embed}

@@

\begin{definition}\label{def:embed2}
  Resolutive and factive predicates:
  \begin{itemize}
  \item \term{Factive predicates} take both, questions and propositions, and
  demand the extension of (the set of) propositions they embed to evaluate to true (exhaustively).
  Examples:
  \example{know}, \example{forget}, \example{reveal}, \example{discover},
  \item \term{Resolutive predicates} take both, questions and propositions, and
  do not demand the extension of (set of propositions) they embed to have any particular truth value.
  Examples:
  \example{report}, \example{predict}
  \end{itemize}
\end{definition}

@@

The underlying theory for 
TODO Discussion of Lewis' paper.

@@

\subsection{Advertent Curt}

@@

\subsection{The Syntax of Embedded Questions}

@@

However, the syntax of embedded interrogatives is a more delicate matter. Since,
as was discussed in section \ref{sec:embed}, not every embedding verb can embed
every type of sentence (recall definition \ref{def:embed1} and definition
\ref{def:embed2}), a type system has to be established in order to account for
the different behaviour of the verbs, and said type system has to be integrated
into the process of syntactic and semantic parsing of the system.

\curt{} parses sentences of natural language using an enhanced version of \pn{Cooper
storage}\footnote{See \cite{cooper:storage2} for the original implementation.} as
presented in \cite{keller:storage}. \cite{blackburnbos:cl1} call this version
\pn{Keller Storage}. The remainder of this section will follow the nomenclature
of \cite{blackburnbos:cl1}.

@@

Enhancing the grammatic rules of the \curt{} system to cope with
proposition-embedding verbs is made easy by its elaborate grammatical back
end. In particular, four files are enhanced with rules to enable interpretation
of embedded questions: \code{englishGrammar.pl}, which holds generic \abbr{DCG}
rules for syntactic parsing, \code{englishLexicon.pl}, a dictionary that uses
feature structures to keep track of agreement, and two \pn{Keller Storage}
specific files, \code{semLexStorage.pl} and \code{semRulesKeller.pl}.

The grammatic rules are extended by the introduction of three new categories,
\code{whemb}, \code{tbar} and \code{ivtbar}.

The \code{whemb} and \code{ivtbar} categories have a \code{type} parameter, that dictates
their behaviour, although their respective type systems never cooperate with
each other. The type of the \code{whemb} category can hold the values
\code{propos}, for a proposition embedding complementizer, like \example{that},
and \code{question}, for a complementizer that turns a proposition into a yes/no
interrogative, like \example{if} and \example{whether} (listing \ref{whemb}).

\begin{lstlisting}[label=whemb,caption={Syntactic and Lexical Rules for Complementizers},float]
whemb([sem:Sem]) -->
    {lexEntry(whemb,[syntax:Word,type:T])}
    , Word
    , {semLex(whemb,[sem:Sem,type:T])}
    .
lexEntry(whemb,[syntax:[that],type:propos]).
lexEntry(whemb,[syntax:[whether],type:question]).
\end{lstlisting}

\begin{lstlisting}[label=whembsemLex,caption={Semantic Rules for Complementizers},float]
semLex(whemb,M) :-
    M = [sem:[lam(S,S)],type:propos]
    .
semLex(whemb,M) :-
    M = [sem:[lam(Q, que([],[],Q))],type:question]
    .
\end{lstlisting}

The types now determine whether to produce an embedded yes/no
interrogative or just pass through as a simple proposition (in which case a
simple $\lambda S. S$ serves as an identity function), and the
\code{semLex/2} predicate takes care of this (listing \ref{whembsemLex}). In the
latter case, a new sort of question is introduced, namely one that does not
quantify over a free variable. It has the form \lstinline!que([],[],Q)! and is
representative of $?.\varphi$.

The \code{tbar} rule has three different right hand sides. 
In one of them, it first parses a \code{whemb} and combines it with a sentential
category \code{s} (listing \ref{tbar}, line 1). The second case omits the
complementizer and just passes the sentence as a normal proposition (line 7).
Next, the final rule takes care of embedding questions (line 12).

In every case, the sentence embedded is first fully reduced by the predicates
shown in listing \ref{tbarCombine}. This has to be done in order to avoid
unsaturated formulae in the end result. Thus the $\bar{t}$ category can   produce
one of the following: $\varphi$, $?.\varphi$ or $?x.\varphi$.

\begin{lstlisting}[label=tbar,caption={Semantic Rules for $\bar{t}$},float]
tbar([sem:TBar]) -->
    whemb([sem:WHemb])
    , s([coord:no,sem:S])
    , { combine(tbar:TBar,[whemb:WHemb,s:S]) } .

tbar([sem:TBar]) -->
    s([coord:no,sem:S])
    , { combine(tbar:TBar,[s:S]) } .

tbar([sem:TBar]) -->
    q([sem:Q])
    , { combine(tbar:TBar,[s:Q]) } .
\end{lstlisting}

\begin{lstlisting}[label=tbarCombine,caption={$\bar{t}$ is responsible for beta-converting the embedded sentence.},float]
combine(tbar:[app(WH,SRed)],[whemb:[WH],s:S]) :-
    betaConvert(S,SRed).

combine(tbar:[SRed],[s:S]) :-
    betaConvert(S,SRed).
\end{lstlisting}

As a final step, \code{ivtbar} (or $IV/\bar{t})$ in a formal type system), takes
the result of \code{tbar} and embeds it under a suitable verb, taking the verb
type (one of \code{factive}, \code{resolutive}, \code{question} and
\code{propos}) and the nature of the embedded sentence into account. Listing
\ref{ivtbar} shows only one case, for \code{question} embedding verbs, and the
(semantic) lexical rules have been omitted. See the
appendix %todo: section number
for the full definition.

\begin{lstlisting}[label=ivtbar,caption={A question embedding VP requires its antecedent to be a question.},float]
vp([coord:no,inf:Inf,num:Num,gap:[],sem:VP]) -->
    ivtbar([inf:Inf,num:Num,sem:IVTBAR,type:question])
   , tbar([sem:TBAR])
   , { TBAR = [app(lam(X,que(_,_,X)),X)]
       , combine(vp:VP,[ivtbar:IVTBAR,tbar:TBAR])
     }.
\end{lstlisting}
